---
title: "Data Science Project 1"
date: "`r Sys.Date()`"
author: 
  - "Constantine Zhang"
  - "Ha Nguyen"
  - "Kemal Ege Sural"
output: 
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(results="markup", warning = F, message = F,
                      fig.width=8, fig.height=6, fig.dim=c(8,6), fig.align='center')
options(scipen = 999, digits = 3) 
```

```{r library, include = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(tinytex)
library(kableExtra)
```

```{r data import and cleaning na value, include = FALSE}
data_counties <- read.csv("introdata_merged_counties.csv")
data_counties %>% drop_na()
```

```{r preview data}
tmp <- head(data_counties, 5)
names(tmp) <- c("County FIPS","State","County",
                "Per-Capita Income (2019)","Per-Capita Income (2020)","Per-Capita Income (2021)",
                "Associate Degrees (Count, 2016–2020)","Bachelor’s Degrees (Count, 2016–2020)",
                "Associate Degrees (%)","Bachelor’s Degrees (%)",
                "Population (Est.)","Population (Male)","Population (Female)",
                "Under 5 (Count)","Age 18+ (Count)","Age 65+ (Count)","Median Age")

kable(tmp, caption = "Table 1. First 5 rows of counties data",
      format = "html", align = "c") |>
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed"),
                            full_width = FALSE, position = "center") |>
  kableExtra::scroll_box(width = "100%", height = "300px")
```

```{r data set summary}
summary(data_counties)
```
```{r}
ggplot(data_counties, aes(x = bachelor_degree_numbers_2016_2020, y = per_capita_personal_income_2020 )) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Bachelor’s Degree Numbers vs. Per Capita Personal Income (2020) Scatter Plot",
    x = "Number of Bachelor's Degree 2016-2020",
    y = "Personal Income Per Capita (2020)"
  ) 

ggplot(data_counties, aes(x = associate_degree_numbers_2016_2020 , y = per_capita_personal_income_2020 )) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Associate’s Degree Numbers vs. Per Capita Personal Income (2020) Scatter Plot",
    x = "Number of Associate's Degree 2016-2020",
    y = "Personal Income Per Capita (2020)"
  ) 

```

```{r scatter plot after exclude extrem value}
ggplot(data_counties, aes(x = bachelor_degree_numbers_2016_2020, y = per_capita_personal_income_2020 )) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Bachelor’s Degree Numbers vs. Per Capita Personal Income (2020) 
    Scatter Plot - Exclude extreme value",
    x = "Number of Bachelor's Degree 2016-2020",
    y = "Personal Income Per Capita (2020)" 
  ) + 
  scale_x_continuous(limits = c(0, 150000))  + 
  scale_y_continuous(limits = c(0, 100000))


ggplot(data_counties, aes(x = associate_degree_numbers_2016_2020 , y = per_capita_personal_income_2020 )) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Associate’s Degree Numbers vs. Per Capita Personal Income (2020) 
    Scatter Plot - Exclude extreme value",
    x = "Number of Associate's Degree 2016-2020",
    y = "Personal Income Per Capita (2020)"
  ) +
  scale_x_continuous(limits = c(0, 150000))  +
  scale_y_continuous(limits = c(0, 100000))

```

```{r exclude extreme value}
# find an metrics for determine extreme values, we exclude extreme value that is more than 5 # standard deviation away
quantile(data_counties$bachelor_degree_numbers_2016_2020, prob =0.95)
quantile(data_counties$associate_degree_numbers_2016_2020, prob =0.95)

```

```{r  finding Confidence Intervals for income, degree, population}
per_capita_2019_ttest <- t.test(x=data_counties$per_capita_personal_income_2019, conf.level=0.95)
per_capita_2019_ci_95 <- per_capita_2019_ttest$conf.int

per_capita_2020_ttest <- t.test(x=data_counties$per_capita_personal_income_2020, conf.level=0.95)
per_capita_2020_ci_95 <- per_capita_2020_ttest$conf.int

per_capita_2021_ttest <- t.test(x=data_counties$per_capita_personal_income_2021, conf.level=0.95)
per_capita_2021_ci_95 <- per_capita_2021_ttest$conf.int

bachelor_ttest <- t.test(x=data_counties$bachelor_degree_percentage_2015_2019, conf.level=0.95)
bachelor_ci_95 <- bachelor_ttest$conf.int

associate_ttest <- t.test(x=data_counties$associate_degree_percentage_2016_2020, conf.level=0.95)
associate_ci_95 <- associate_ttest$conf.int

population_ttest <- t.test(x=data_counties$POPESTIMATE, conf.level=0.95)
population_ci_95 <- population_ttest$conf.int
```

```{r finding Confidence interval for age}
median_age_ttest <- t.test(x=data_counties$MEDIAN_AGE_TOT, conf.level=0.95)
median_age_ci_95 <- median_age_ttest$conf.int

z        <- qnorm(0.975)
sum_pop <- sum(data_counties$POPESTIMATE,na.rm = TRUE)
share5 <- sum(data_counties$UNDER5_TOT,na.rm = TRUE) / sum_pop
se5 <- sqrt(share5 * (1 - share5) / sum_pop)
share5_ci_95 <- c(share5 - z*se5, share5 + z*se5)
share5_ci_95_pct <- share5_ci_95 * 100

share65 <- sum(data_counties$AGE65PLUS_TOT,na.rm = TRUE) / sum_pop
se65 <- sqrt(share65 * (1 - share65) / sum_pop)
share65_ci_95 <- c(share65 - z*se65, share65 + z*se65)
share65_ci_95_pct <- share65_ci_95 * 100
```

**Per-capita income (2019–2021)**:

- ***The mean county per-capita income in 2019*** is estimated to lie between `r scales::comma(round(per_capita_2019_ci_95[1], 2))` and `r scales::comma(round(per_capita_2019_ci_95[2], 2))` (95% CI).

- ***For 2020***, the 95% CI is `r scales::comma(round(per_capita_2020_ci_95[1], 2))` to `r scales::comma(round(per_capita_2020_ci_95[2], 2))`

- ***For 2021***, it is `r scales::comma(round(per_capita_2021_ci_95[1], 2))` to `r scales::comma(round(per_capita_2021_ci_95[2], 2))`.

The limited overlap across years indicates a statistically meaningful upward shift in the average county income.

**Population (mean per county)**:

- ***The mean county population*** is between `r scales::comma(round(population_ci_95[1], 2))` and `r scales::comma(round(population_ci_95[2], 2))` (95% CI). This is a wide band that reflects substantial heterogeneity across counties (very small rural counties vs large urban ones). This reflects the average county, not the national total. 

**Education**:

- ***Bachelor’s share*** (mean across counties): `r if (exists("bachelor_ci_95")) paste0("[", round(bachelor_ci_95[1],2), ", ", round(bachelor_ci_95[2],2), "]")`.

- ***Associate share *** (mean across counties): `r if (exists("associate_ci_95")) paste0("[", round(associate_ci_95[1],2), ", ", round(associate_ci_95[2],2), "]")`.

**Population composition (person-weighted) by age**:

- ***Median age*** (mean of county medians): The mean of county median ages lies between `r round(median_age_ci_95[1], 2)` and `r round(median_age_ci_95[2], 2)` (95% CI), characterizing the average county’s age profile. As we can see here, age profile is tight and middle-aged.


- ***Age 65+ overall share***: In the pooled population, the estimated share age 65+ is `r scales::percent(share65, accuracy = 0.01)`, with a 95% CI of `r scales::percent(share65_ci_95[1], accuracy = 0.01)` to `r scales::percent(share65_ci_95[2], accuracy = 0.01)`. This is a population-weighted estimate (ratio of totals), so each person counts equally.

- ***Under 5 overall share***: The overall share under age 5 is `r scales::percent(share5, accuracy = 0.01)`, with a 95% CI of `r scales::percent(share5_ci_95[1], accuracy = 0.01)` to `r scales::percent(share5_ci_95[2], accuracy = 0.01)`.

```{r  T-test to compare mean of income and degree between counties}
#In this section, we will conduct annova test to test the null hypothesis of difference in mean of income and degree across states
```

```{r  Chi-squares}
#In this section, we will conduct annova test to test the null hypothesis of relationship between income and degree/ gender/ age across states
```





