---
title: "Data Science Project 1"
date: "`r Sys.Date()`"
author: 
  - "Constantine Zhang"
  - "Ha Nguyen"
  - "Kemal Ege Sural"
output: 
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(results="markup", warning = F, message = F,
                      fig.width=8, fig.height=6, fig.dim=c(8,6), fig.align='center')
<<<<<<< HEAD
options(scipen = 999, digits = 3) 
=======
options(scientific=T, digits = 3,scipen = 999) 
>>>>>>> 2b8d385b3cbde474d3aecb1f523d883d3e7a5da2
```

```{r library, include = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(tinytex)
library(kableExtra)
```

```{r data import and cleaning na value, include = FALSE}
data_counties <- read.csv("introdata_merged_counties.csv")
data_counties %>% drop_na()
```

```{r preview data}
tmp <- head(data_counties, 5)
names(tmp) <- c("County FIPS","State","County",
                "Per-Capita Income (2019)","Per-Capita Income (2020)","Per-Capita Income (2021)",
                "Associate Degrees (Count, 2016–2020)","Bachelor’s Degrees (Count, 2016–2020)",
                "Associate Degrees (%)","Bachelor’s Degrees (%)",
                "Population (Est.)","Population (Male)","Population (Female)",
                "Under 5 (Count)","Age 18+ (Count)","Age 65+ (Count)","Median Age")

kable(tmp, caption = "Table 1. First 5 rows of counties data",
      format = "html", align = "c") |>
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed"),
                            full_width = FALSE, position = "center") |>
  kableExtra::scroll_box(width = "100%", height = "300px")
```

```{r data set summary}
summary(data_counties)
```

```{r}
ggplot(data_counties, aes(x = bachelor_degree_numbers_2016_2020, y = per_capita_personal_income_2020 )) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Bachelor’s Degree Numbers vs. Per Capita Personal Income (2020) Scatter Plot",
    x = "Number of Bachelor's Degree 2016-2020",
    y = "Personal Income Per Capita (2020)"
  ) 

ggplot(data_counties, aes(x = associate_degree_numbers_2016_2020 , y = per_capita_personal_income_2020 )) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Associate’s Degree Numbers vs. Per Capita Personal Income (2020) Scatter Plot",
    x = "Number of Associate's Degree 2016-2020",
    y = "Personal Income Per Capita (2020)"
  ) 

```

```{r scatter plot after exclude extrem value}
ggplot(data_counties, aes(x = bachelor_degree_numbers_2016_2020, y = per_capita_personal_income_2020 )) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Bachelor’s Degree Numbers vs. Per Capita Personal Income (2020) 
    Scatter Plot - Exclude extreme value",
    x = "Number of Bachelor's Degree 2016-2020",
    y = "Personal Income Per Capita (2020)" 
  ) + 
  scale_x_continuous(limits = c(0, 150000))  + 
  scale_y_continuous(limits = c(0, 100000))


ggplot(data_counties, aes(x = associate_degree_numbers_2016_2020 , y = per_capita_personal_income_2020 )) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Associate’s Degree Numbers vs. Per Capita Personal Income (2020) 
    Scatter Plot - Exclude extreme value",
    x = "Number of Associate's Degree 2016-2020",
    y = "Personal Income Per Capita (2020)"
  ) +
  scale_x_continuous(limits = c(0, 150000))  +
  scale_y_continuous(limits = c(0, 100000))

```

```{r exclude extreme value}
# find an metrics for determine extreme values, we exclude extreme value that is more than 5 # standard deviation away
quantile(data_counties$bachelor_degree_numbers_2016_2020, prob =0.95)
quantile(data_counties$associate_degree_numbers_2016_2020, prob =0.95)

```

```{r  finding Confidence Intervals for income, degree, population}
per_capita_2019_ttest <- t.test(x=data_counties$per_capita_personal_income_2019, conf.level=0.95)
per_capita_2019_ci_95 <- per_capita_2019_ttest$conf.int

per_capita_2020_ttest <- t.test(x=data_counties$per_capita_personal_income_2020, conf.level=0.95)
per_capita_2020_ci_95 <- per_capita_2020_ttest$conf.int

per_capita_2021_ttest <- t.test(x=data_counties$per_capita_personal_income_2021, conf.level=0.95)
per_capita_2021_ci_95 <- per_capita_2021_ttest$conf.int

bachelor_ttest <- t.test(x=data_counties$bachelor_degree_percentage_2015_2019, conf.level=0.95)
bachelor_ci_95 <- bachelor_ttest$conf.int

associate_ttest <- t.test(x=data_counties$associate_degree_percentage_2016_2020, conf.level=0.95)
associate_ci_95 <- associate_ttest$conf.int

population_ttest <- t.test(x=data_counties$POPESTIMATE, conf.level=0.95)
population_ci_95 <- population_ttest$conf.int
```

```{r finding Confidence interval for age}
median_age_ttest <- t.test(x=data_counties$MEDIAN_AGE_TOT, conf.level=0.95)
median_age_ci_95 <- median_age_ttest$conf.int

z        <- qnorm(0.975)
sum_pop <- sum(data_counties$POPESTIMATE,na.rm = TRUE)
share5 <- sum(data_counties$UNDER5_TOT,na.rm = TRUE) / sum_pop
se5 <- sqrt(share5 * (1 - share5) / sum_pop)
share5_ci_95 <- c(share5 - z*se5, share5 + z*se5)
share5_ci_95_pct <- share5_ci_95 * 100

share65 <- sum(data_counties$AGE65PLUS_TOT,na.rm = TRUE) / sum_pop
se65 <- sqrt(share65 * (1 - share65) / sum_pop)
share65_ci_95 <- c(share65 - z*se65, share65 + z*se65)
share65_ci_95_pct <- share65_ci_95 * 100
```

**Per-capita income (2019–2021)**:

- ***The mean county per-capita income in 2019*** is estimated to lie between `r scales::comma(round(per_capita_2019_ci_95[1], 2))` and `r scales::comma(round(per_capita_2019_ci_95[2], 2))` (95% CI).

- ***For 2020***, the 95% CI is `r scales::comma(round(per_capita_2020_ci_95[1], 2))` to `r scales::comma(round(per_capita_2020_ci_95[2], 2))`

- ***For 2021***, it is `r scales::comma(round(per_capita_2021_ci_95[1], 2))` to `r scales::comma(round(per_capita_2021_ci_95[2], 2))`.

The limited overlap across years indicates a statistically meaningful upward shift in the average county income.

**Population (mean per county)**:

- ***The mean county population*** is between `r scales::comma(round(population_ci_95[1], 2))` and `r scales::comma(round(population_ci_95[2], 2))` (95% CI). This is a wide band that reflects substantial heterogeneity across counties (very small rural counties vs large urban ones). This reflects the average county, not the national total. 

**Education**:

- ***Bachelor’s share*** (mean across counties): `r if (exists("bachelor_ci_95")) paste0("[", round(bachelor_ci_95[1],2), ", ", round(bachelor_ci_95[2],2), "]")`.

- ***Associate share *** (mean across counties): `r if (exists("associate_ci_95")) paste0("[", round(associate_ci_95[1],2), ", ", round(associate_ci_95[2],2), "]")`.

**Population composition (person-weighted) by age**:

- ***Median age*** (mean of county medians): The mean of county median ages lies between `r round(median_age_ci_95[1], 2)` and `r round(median_age_ci_95[2], 2)` (95% CI), characterizing the average county’s age profile. As we can see here, age profile is tight and middle-aged.


- ***Age 65+ overall share***: In the pooled population, the estimated share age 65+ is `r scales::percent(share65, accuracy = 0.01)`, with a 95% CI of `r scales::percent(share65_ci_95[1], accuracy = 0.01)` to `r scales::percent(share65_ci_95[2], accuracy = 0.01)`. This is a population-weighted estimate (ratio of totals), so each person counts equally.

- ***Under 5 overall share***: The overall share under age 5 is `r scales::percent(share5, accuracy = 0.01)`, with a 95% CI of `r scales::percent(share5_ci_95[1], accuracy = 0.01)` to `r scales::percent(share5_ci_95[2], accuracy = 0.01)`.

```{r  T-test to compare mean of income and degree between counties}
#In this section, we will conduct annova test to test the null hypothesis of difference in mean of income and degree across states
data_counties$state <- as.factor(data_counties$state)
anova_model_Income <- aov(per_capita_personal_income_2020 ~ state, data = data_counties)
summary(anova_model_Income)
#group by to get state level mean
state_means <- data_counties %>%
  group_by(state) %>%
  summarize(mean_income = mean(per_capita_personal_income_2020, na.rm = TRUE))
#bar chart to look at distribution of state level mean income 
ggplot(state_means, aes(x = state, y = mean_income)) +
  geom_col(fill = "skyblue") +
  labs(title = "Mean Income by State",
       x = "State",
       y = "Mean Income") +
  theme_minimal()
```
This result is statistically significant at even 1% level. That imply there is a difference in average income between states at 1% level.  

```{r}
anova_model_Bachelor <- aov(bachelor_degree_numbers_2016_2020 ~ state, data = data_counties)
summary(anova_model_Bachelor)
#group by to get state level mean
state_means_bachelor <- data_counties %>%
  group_by(state) %>%
  summarize(bachelor_degree_numbers = mean(bachelor_degree_numbers_2016_2020, na.rm = TRUE))
#bar chart to look at distribution of state level bachelor number 
ggplot(state_means_bachelor, aes(x = state, y = bachelor_degree_numbers )) +
  geom_col(fill = "skyblue") +
  labs(title = "Mean Income by State",
       x = "State",
       y = " bachelor degree numbers") +
  theme_minimal()
```



```{r}
anova_model_Associate <- aov(associate_degree_numbers_2016_2020 ~ state, data = data_counties)
summary(anova_model_Associate)
#group by to get state level mean 
state_means_associate <- data_counties %>%
  group_by(state) %>%
  summarize(associate_degree_numbers = mean(associate_degree_numbers_2016_2020, na.rm = TRUE))
#bar chart to look at distribution of state level bachelor number 
ggplot(state_means_associate, aes(x = state, y = associate_degree_numbers )) +
  geom_col(fill = "skyblue") +
  labs(title = "Mean Income by State",
       x = "State",
       y = " associate degree numbers") +
  theme_minimal()
```


```{r  Chi-squares}
#In this section, we will conduct annova test to test the null hypothesis of relationship between income and degree/ gender/ age across states
```

```{r Per Capita Personal Income Distribution by State (2020)}

data_counties_filtered <- data_counties %>%
  group_by(state) %>%
  mutate(median_income_state = median(per_capita_personal_income_2020, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(state %in% unique(.$state)[1:15]) 

ggplot(data_counties_filtered, aes(x = reorder(state, per_capita_personal_income_2020, FUN = median), 
                                   y = per_capita_personal_income_2020, 
                                   fill = state)) +
  geom_boxplot() +
  coord_flip() + 
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Per Capita Personal Income Distribution by State (2020)",
    x = "State",
    y = "Per Capita Personal Income (2020)"
  )
```

The box plot for "Per Capita Personal Income Distribution by State (2020)" provides a summary assessment of county level economic differences within the sampled states. The visualization focuses on the top 15 states, which were selected to ensure the clarity and readability of the detailed comparisons. The analysis of central tendency indicates clear economic tiers: Massachusetts (MA) exhibits the highest median per capita income, placing it alongside New Jersey (NJ) and Maryland (MD) in the most affluent group. Conversely, states like Tennessee (TN) display a distribution concentrated at the lower end of the income scale. Colorado (CO) and Ohio (OH) present the widest boxes, signifying the greatest income disparity in income across their counties. Finally, the extreme high outliers observed in New York (NY) and California (CA) highlight the substantial concentration of wealth in a limited number of exceptional counties that significantly influence state averages.





